Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-04-27 manuscript/Manuscript.Rmd
Edited MS

# Introduction
Women are substantially underrepresented in the workforce in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and they face additional challenges and inequities relative to men [e.g. @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. For example, women tend to occupy more junior positions (Wutte 2007, Reuben, Sapienza & Zingales, 2014) with lower salaries (Trower and Chait 2002, Umback 2007), receive less grant money (Hosek et al. 2005, OER 2005), are promoted more slowly (Hopkins et al. 2013; Long et al. 1993, Rosenfeld 1991, Zuckerman 1987), and are allocated fewer resources (European Commission 2009) and less research funding (Feldt 1986, Stack 2004, Larivière et al. 2011) than men. Experimental studies have demonstrated that researchers regard women's achievements less favourably than identical achievements by a man (Moss et al. 2012; Knobloch et al. 2013). Multiple studies have found that women publish fewer papers per year than men (Cole and Zuckerman 1984, Long 1992, Larivière et al. 2013, Astronomy one in Nature) [check whether they control for career stage].

@@

A high, steadily increasing proportion of research papers is written by more than one author (West et al. 2013), making collaboration a key predictor of publication output, and in turn, a key predictor of career prospects (Cohn et al 2014, Jordan et al. 2008, Tower et al. 2007). Previous studies suggest that mixed-gender (Campbell et al. 2013) or otherwise diverse (Hong and Page 2004) teams tend to produce better results on collaborative tasks. For these reasons, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every study of this question has concluded that men publish with other men, and women with women, more often than expected if collaborators assort randomly with regards to gender (Ferber and Teiman, 1980; McDowell and Smith 1992; Ghiasi et al. 2015; Crow and Smykla 2015, Zettler et al. 2016, Fahmy and Young, 2016, Jadidi et al. 2017; Teele and Thelen 2017, Araujo and Fontainha 2017a,b). This phenomenon is often termed ‘gender homophily’.

Unfortunately, these results all derive from studies that used sampling strategies that could overstate the strength of homophily, making the true extent of homophily difficult to ascertain from present data (\autoref{wahlund_plot}). Essentially, whenever samples are taken from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can make it appear as though authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers prefer opposite-gendered colleagues. For example, a sample of bioinformatics and cell biology papers might contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because more women working in cell biology than in bioinformatics [@Holman_2018]. We name this issue the Wahlund effect, by analogy with the Wahlund effect in population genetics (Wahlund 1928), whereby spatial differences in allele frequencies create an excess of homozygotes, even if alleles are randomly assorted within each subpopulation.

@@

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, even after implementing stringent controls for ‘spurious’ homophily resulting from Wahlund effects (Figure 1). Our data therefore concur with previous conclusions (Ferber and Teiman 1980; McDowell and Smith 1992; Crow and Smykla 2015; Ghiasi et al. 2015; Zettler et al. 2016; Fahmy and Young 2016; Jadihi et al. 2017; Araujo and Fontainha 2017a; Teele and Thelen 2017) that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Moreover, we found that gender homophily is not unique to particular disciplines, but rather appears to be similar in strength across many diverse fields of STEMM.

>>>>

# Introduction
Women are substantially underrepresented in the workforce in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and they face additional challenges and inequities relative to men [e.g. @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions (Wutte 2007, Reuben, Sapienza & Zingales, 2014) with lower salaries (Trower and Chait 2002, Umback 2007), receive less grant money (Hosek et al. 2005, OER 2005), are promoted more slowly (Hopkins et al. 2013; Long et al. 1993, Rosenfeld 1991, Zuckerman 1987), and are allocated fewer resources (European Commission 2009) and less research funding (Feldt 1986, Stack 2004, Larivière et al. 2011) than men. Experimental studies have demonstrated that researchers regard women's achievements less favourably than identical achievements by men (Moss et al. 2012; Knobloch et al. 2013). Multiple studies have found that women publish fewer papers per year than men (Cole and Zuckerman 1984, Long 1992, Larivière et al. 2013, Astronomy one in Nature) [check whether they control for career stage].

@@

A high, steadily increasing proportion of research papers is written by more than one author (West et al. 2013), making collaboration a key predictor of publication output, and in turn, a key predictor of career prospects (Cohn et al 2014, Jordan et al. 2008, Tower et al. 2007). Additionally, empirical studies suggest that mixed-gender (Campbell et al. 2013) or otherwise diverse (Hong and Page 2004) teams can produce better results on collaborative tasks. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every study of this question has concluded that men publish with other men, and women with women, more often than expected if collaborators assort randomly with regards to gender (Ferber and Teiman, 1980; McDowell and Smith 1992; Ghiasi et al. 2015; Crow and Smykla 2015, Zettler et al. 2016, Fahmy and Young, 2016, Jadidi et al. 2017; Teele and Thelen 2017, Araujo and Fontainha 2017a,b). This preference for same-gendered colleagues is often termed ‘gender homophily’.

However, we believe that all of these studies used sampling strategies that could overstate the strength of homophily, making the true extent of homophily difficult to ascertain from present data. Essentially, whenever samples are taken from two or more discrete sets of literature that vary in the author gender ratio and that are largely not connected by collaboration, the number of same-gendered coauthors will be inflated (\autoref{wahlund_plot}). This can make it appear as though authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers prefer opposite-gendered colleagues. For example, a sample of bioinformatics and cell biology papers might contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because more women work in cell biology than in bioinformatics [@Holman_2018]. We name this issue the Wahlund effect, by analogy with the Wahlund effect in population genetics (Wahlund 1928), whereby spatial differences in allele frequencies create an excess of homozygotes across the whole population, even if alleles assort randomly within each subpopulation.

@@

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, even after implementing stringent controls for ‘spurious’ homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore concur with previous conclusions (Ferber and Teiman 1980; McDowell and Smith 1992; Crow and Smykla 2015; Ghiasi et al. 2015; Zettler et al. 2016; Fahmy and Young 2016; Jadihi et al. 2017; Araujo and Fontainha 2017a; Teele and Thelen 2017) that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Moreover, we found that gender homophily is not unique to particular disciplines, but rather appears to be similarly strong across many diverse fields of STEMM.
